\Title

### 2013-02-13 synmut.tex
text small changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\timedependence}{4}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. Synonymous mutations do not affect 
the interaction with the adaptive immune system and are often
assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion{} are only transiently beneficial, either
because the immune system is catching up or because of competition between equivalent
escapes.

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which change to avoid recognition
by neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
In addition to maintaining protein function and avoiding the adaptive immune
recognition, however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes since these processes often
depend on RNA folding.
For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.

Even in absense of important RNA structure synonymous codons don't evolve
completely neutrally but that some codons are favored over others in many
species \citep{plotkin_synonymous_2011}.  Recent studies have shown that
genetically engineered HIV strains with altered codon usage can in some cases
produce more viral protein, but in general replicate less
efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.  Codon-deoptimization has been
suggested as attenuation strategy for polio and
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying selection
beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
rates of evolution at synonymous sites vary along the HIV genome
\citep{mayrose_towards_2007}.  Positive selection through the host's immune
system, however, is restricted to changes in the amino acid sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(i) exactly one individual in the current population will be
the common ancestor of the entire future population at a particular locus and
(ii) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft} shows 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env, C2-V5, in patient p10~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, 
few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (patients whose viral population was
structured were excluded from the analysis; see methods).
The former data set from is restricted to the
C2-V5 region of \env, while the data from \citet{bunnik_autologous_2008} covers
the majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t+\Delta t$. Plotting these fixed, lost, and polymorphic fraction
against the time interval $\Delta t$, we see that most synonymous mutations
segregate for roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
gotten to high frequency by chance (genetic drift or hitchhiking), and
we expect that selection will drive it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env~that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations in this
region are neutral -- only those mutations observed at high frequencies, which
have been experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is reduced relative to the neutral
expectation indicated by the diagonal line. This suppression is not
observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous
mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
which is deleterious to the virus
~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011,sanjuan_interplay_2011}.

The propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles that are lost
(blue ) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that are not observed (green) lies between the
those that fix and those that get lost.
Note that, that this analysis will be sensitive only at position where the base
pairing pattern of NL4-3 agrees with that of each patient's initial
consensus sequence.

To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved stretches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles consistent with that HIV does not seem to adapt its codon
usage to its human host cells at the macroevolutionary level
\citep{kuyl_biased_2012}.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
that fix, mutations that reach frequencies above 15\%, but
are susequently lost, and random mutations (all categories are restricted to the
regions V1-V5$\pm 100$bp).
Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
less base pairing ($p=0.002$, Kolmogorov-Smirnov two sample test on the
corresponding cumulative distribution).
Mutations that are never observered show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions that
harbor conserved RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. When killer T cells or antibodies
target a new epitope, the virus often escapes via nonsynonymous mutations that
impair recognition. Synonymous mutations do not affect this interplay and are
often assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are
brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles. The patterns of fixation of nonsynonymous mutations
suggest that escape mutations in \shankaregion{} are only transiently
beneficial, either because the immune system is catching up or because of
competition between equivalent escapes.

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate of
HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
neutralizing ABs. Mutations in \env, the gene encoding gp120, spread through the
population within a few months (see \figurename~\ref{fig:aftnonsyn}).
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 month earlier
\citep{richman_rapid_2003}.

These escape mutations are selected for their effect on the amino acid sequence
of the viral proteins. Conversely, synonymous mutations are commonly used as
approximately neutral markers in studies of viral evolution. Neutral markers are
very useful since their dynamics can be compared to that of putatively
functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes since these processes
often depend on RNA folding. For example, the HIV \rev{} response element (RRE)
in \env{} enhances nuclear export of full length or partially spliced viral
transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse
transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
required for priming reverse transcription (RT) and is bound by a pseudoknotted
RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}.

Even in absence of important RNA structures, synonymous codons do not evolve
completely neutrally. Some codons are favored over others in many species
\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
engineered HIV strains with altered codon usage can in some cases produce more
viral protein, but in general replicate less efficiently
\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
keating_rich_2009}. Codon deoptimization has been suggested as attenuation
strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
Purifying selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
rates of evolution at synonymous sites vary along the HIV genome
\citep{mayrose_towards_2007}. Positive selection through the host adaptive
immune system, however, is restricted to changes in the amino acid sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency. A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. As illustrated in the inset of
\FIG{aftsyn}, this is a simple consequence of the fact that (i) exactly one
individual in the current population will be the common ancestor of the entire
future population at a particular locus and (ii) this ancestor has a probability
$\nu$ of carrying the mutation (assuming the neutral mutation is not
preferentially associated with genomes of high or low fitness). Deleterious or
beneficial mutations fix less or more often than neutral ones, respectively.
\FIG{aft} shows the time course the frequencies of all synonymous and
nonsynonymous mutations observed \env, \shankaregion, in patient
p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (patients whose viral population was structured
were excluded from the analysis; see methods and \figurename~S\PCApat).  The
former data set from is restricted to the \shankaregion{} region of \env, while
the data from \citet{bunnik_autologous_2008} covers the majority of \env.
Considering all mutations in a frequency interval around $\nu_0$ at some time
$t$, we calculate the fraction that is found at frequency 1, at frequency 0, or
at intermediate frequency at later times $t+\Delta t$. Plotting these fixed,
lost, and polymorphic fraction against the time interval $\Delta t$, we see that
most synonymous mutations segregate for roughly one year and are lost much more
frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
fixation versus extinction of synonymous mutations is shown as a function of the
initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). Restricted to the
region \shankaregion, we find that $\pfix$ of synonymous mutations is far below
the neutral expectation.  Outside of \shankaregion, using data from
\citet{bunnik_autologous_2008} only, no such reduction in $\pfix$ is found.
Restricted to the \shankaregion area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will
come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the \shankaregion region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

When interpreting these results for the fixation probabilities, it is important
to distinguish between random mutations and polymorphisms observed at a certain
frequency since the latter have already been filtered by selection.
A polymorphism could be beneficial to the virus and on its way to fixation. In
this case, we expect that it fixes almost surely given we see it at high
frequency. If, on the other hand, the polymorphism is deleterious it must have
reached a high frequency by chance (genetic drift or hitchhiking), and
we expect that selection drives it out of the population again. Hence our
observations suggest that many of the synonymous polymorphisms at intermediate
frequencies in the part of \env{} that includes the hypervariable regions are
deleterious, while outside this regions polymorphisms are mostly roughly
neutral. Note that this does not imply that all synonymous mutations in this
region are neutral -- only those mutations observed at high frequencies, which
have been experiencing selection for some time already, tend to be neutral.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in \shankaregion is reduced relative to the neutral
expectation indicated by the diagonal line. This suppression is not
observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in \shankaregion tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous mutations in
\shankaregion are secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We partition all synonymous alleles observed at
intermediate frequencies above 15\% depending on their final destiny (fixation
or extinction). Subsequently, we align our sequences to the reference NL4-3
strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
reactivities. As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $P\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between the those that fix and
those that get lost. Note that this analysis will be sensitive only at position
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

To test the hypothesis that mutations in \shankaregion{} are lost because they
break stems in the conserved stretches between the variable loops, we consider
mutations in variable loops and conserved parts separately. The biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is consistent with important stem structures in conservered
regions between loops.

In addition to RNA secondary structure, we have considered other possible
explanations for a fitness defect of synonymous mutations, in particular codon
usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
however, any evidence for a contribution of average CUB to the ultimate fate of
synonymous alleles; consistently, HIV does not seem to adapt its codon usage to
its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
that fix, mutations that reach frequencies above 15\%, but
are susequently lost, and random mutations (all categories are restricted to the
regions V1-V5$\pm 100$bp).
Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
less base pairing.
Mutations that are never observered show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in
\shankaregion{} separately for variable loops and the connecting conserved regions that
harbor RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
